---
title: "Rowely_May7"
author: "Brock Rowley"
date: "5/7/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lme4)
library(performance)
library(lubridate)
library(sundry)
library(nlme)

cnlsy <- read_csv(here::here("data", "cnlsy.csv"))
wages <- read_csv(here::here("data", "wages.csv"))
sim_d <- read_csv(here::here("data", "curvilinear-sim.csv"))
d <- read_csv(here::here("data", "longitudinal-sim.csv"))
```

```{r}
m1a <- lmer(g5_spring ~ g4_spring + g3_spring + 
              (1|scid) + (1|distid),
            data = d)

m1b <- lmer(g5_spring ~ g4_spring + g3_spring + (1|distid/scid),
            data = d)

m1c <- lmer(g5_spring ~ g4_spring + g3_spring + 
              (1|distid) + (1|distid:scid),
            data = d)

```

```{r}
d <- d %>%
  group_by(scid) %>%
  mutate(sch_mean_start = mean(g3_fall)) %>%
  group_by(distid) %>%
  mutate(dist_mean_start = mean(g3_fall))

m2 <- lmer(g5_spring ~ g4_spring + g3_spring + 
             sch_mean_start + 
             (g3_spring|scid) + (1|distid),
          data = d)
extract_eq(m2, font_size = "small")

m3 <- lmer(g5_spring ~ g4_spring + g3_spring + 
             sch_mean_start + 
             (g4_spring + g3_spring|scid) + (g4_spring + g3_spring|distid),
          data = d)
extract_eq(m3, font_size = "small")

m4 <- lmer(g5_spring ~ g4_spring + g3_spring + 
             sch_mean_start + sch_mean_start:g3_spring +
             (g4_spring + g3_spring|scid) + 
             (g4_spring + g3_spring|distid),
          data = d)
extract_eq(m4, font_size = "small")

m5 <- lmer(g5_spring ~ g4_spring + g3_spring + 
             dist_mean_start +
             (g4_spring + g3_spring||scid) + (g4_spring + g3_spring||distid),
          data = d)
extract_eq(m5, font_size = "small")

# double bar || forces all the off diagonals to be zero
```

```{r}
l <- d %>% 
  pivot_longer(
    cols = starts_with("g"),
    names_to = "timepoint",
    values_to = "score"
  )
l

wave_frame <- tibble(
  timepoint = paste0(
    "g", 
    rep(3:5, each = 3), 
    rep(c("_fall", "_winter", "_spring"), 3)
  ),
  wave = 0:8
)
wave_frame

l <- left_join(l, wave_frame)
l

m6 <- lmer(score ~ wave +
             (wave|sid) + (wave|scid) + (1|distid),
          data = l)
extract_eq(m6)

m7 <- lmer(score ~ wave + sch_mean_start + dist_mean_start +
             wave:sch_mean_start + wave:dist_mean_start +
             (wave|sid) + (wave|scid) + (sch_mean_start|distid),
          data = l)
extract_eq(m7, font_size = "scriptsize")

m8 <- lmer(score ~ wave * sch_mean_start + dist_mean_start +
             (wave|sid),
          data = l)
extract_eq(m8, font_size = "scriptsize")
```

```{r model_fit}
cnlsy <- cnlsy %>%
  mutate(wave_c = wave - 1) # average score at first timepoint, needs to be zero (0)

m_wave <- lmer(piat ~ wave_c + (wave_c|id),
               data = cnlsy)

arm::display(m_wave)
# Note that each wave is tied to a specific age group (the approximate age of participants at that age).

m_agegrp <- lmer(piat ~ agegrp + (agegrp|id),
                 data = cnlsy,
                 control = lmerControl(optimizer = "bobyqa"))

arm::display(m_agegrp)
# Not possible because it predicts students at zero years old
```

```{r centering}
cnlsy <- cnlsy %>%
  mutate(agegrp_c = agegrp - 6.5) # average score at 6.5 years old
m_agegrp2 <- lmer(piat ~ agegrp_c + (agegrp_c|id),
                 data = cnlsy,
                 control = lmerControl(optimizer = "bobyqa"))

arm::display(m_agegrp2)
# What does the intercept represent now?
# Students at 6.5 years old
```

```{r comparing_fit}
compare_performance(m_agegrp, m_agegrp2) %>%
  print_md()
# Table: Comparison of Model Performance Indices
# They are identical!

pred_frame <- tibble(
  agegrp = 0:12,
  agegrp_c = 0:12,
  id = -999
  ) %>%  
  mutate(pred_raw = predict(m_agegrp, newdata = ., allow.new.levels = TRUE),
         pred_c = predict(m_agegrp2, newdata = ., allow.new.levels = TRUE))

points <- data.frame(
  pred_raw = c(fixef(m_agegrp)[1], fixef(m_agegrp2)[1]),
  agegrp = c(0, 6.5),
  color = c("cyan", "magenta")
)

ggplot(pred_frame, aes(agegrp, pred_raw)) +
  geom_line(color = "cornflowerblue",
            size = 5) +
  geom_line(aes(agegrp_c + 6.5, y = pred_c),
            color = "firebrick",
            size = 2) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  geom_segment(x = 0, xend = 0, 
               y = 0, yend = fixef(m_agegrp)[1],
               color = "cyan",
               linetype = "dashed") +
  geom_segment(x = 0, xend = 6.5, 
               y = fixef(m_agegrp2)[1], yend = fixef(m_agegrp2)[1],
               color = "magenta",
               linetype = "dashed") +
  geom_segment(x = 6.5, xend = 6.5, 
               y = 0, yend = fixef(m_agegrp2)[1],
               color = "magenta",
               linetype = "dashed") +
  geom_point(aes(color = color), points, size = 4) +
  guides(color = "none") +
  scale_color_identity() +
  xlim(-5, 20) +
  ylim(-15, 85) +
  labs(x = "Age", y = "Prediction")
```

```{r plot}
# Notice that agegrp does not always correspond directly with their actual age.

ggplot(cnlsy, aes(agegrp, age)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1,
              color = "cornflowerblue")
# Model assumptions:
# When we use the agegrp variable, we are assuming that all children are the exact same age at each assessment wave.
# Although agegrp is more interpretable than wave, it doesn't solve all our problems
```

```{r intercept}
cnlsy %>%
  filter(wave == 1) %>%
  count(age)

# Centering
# I'll choose to subtract 6 from each age
# what will this value represent for students who were 6.91 years old at the first wave?
# Backwards projection

cnlsy <- cnlsy %>%
  mutate(age6 = age - 6)
m_age <- lmer(piat ~ age6 + (age6|id), data = cnlsy)
arm::display(m_age)
```

```{r compare_fit}
compare_performance(m_wave, m_agegrp2, m_age) %>%
  print_md()
# Table: Comparison of Model Performance Indices

# Difference in predictions
pred_frame <- cnlsy %>%
  mutate(pred_agegrp = predict(m_agegrp2),
         pred_age = predict(m_age)) %>%
  filter(id %in% 1:6)

ggplot(pred_frame, aes(age, piat)) +
  geom_point() +
  geom_line(aes(x = age6 + 6, y = pred_age),
            color = "cornflowerblue") +
  geom_line(aes(x = agegrp_c + 6.5, y = pred_agegrp),
            color = "firebrick") +
  facet_wrap(~id)
```

```{r refit}
# Multiply age by 12 to get it coded in months.
cnlsy <- cnlsy %>%
  mutate(age_months = age6 * 12)

m_months <- lmer(piat ~ age_months + (age_months|id), data = cnlsy,
                 control = lmerControl(optimizer = "bobyqa"))
arm::display(m_months)

# Which model fits better? They are not actually the same
compare_performance(m_age, m_months)
# But they are essentially the same
pred_frame %>%
  mutate(pred_months = predict(m_months)[1:18]) %>%
  select(id, starts_with("pred"))
```

## Variables
### id: Participant ID
### lnw: Natural log of wages
### exper: Experience, in years
### ged: Whether or not they completed a GED
### black, hispanic: Dummy variables for race/ethnicity
### hgc: Highest grade completed
### uerate: Unemployment rate at the time

```{r wages_data}
wages %>%
  filter(id %in% c(206, 332))

# Complications = unbalanced data
wages %>%
  count(id) %>%
  summarize(range = range(n))

# Participants age ranged from 14-17 at first time point
# Unequal spacing between waves
```

```{r exper}
m_wage0 <- lmer(lnw ~ exper + (exper|id), data = wages,
                control = lmerControl(optimizer = "bobyqa"))
arm::display(m_wage0)

# Every one year of extra experience corresponded to a 0.05 increase in log wages, on average, which varied across participants with a standard deviation of 0.04.
```

```{r centering_chanllenge}
# Let's center highest grade completed. You could choose whatever value makes the most sense to you. I'll choose Grade 9.

wages <- wages %>%
  mutate(hgc_9 = hgc - 9)

# Is this right? If not, what is it missing?
m_wage1 <- lmer(lnw ~ exper + black + hispanic + hgc_9 +
                  (exper|id),
                data = wages,
                control = lmerControl(optimizer = "bobyqa"))

# Random effects
# In the previous model, we specified exper as randomly varying across id levels.
```

```{r marginal_predictions}
pred_frame <- expand.grid(
  exper = 0:15,
  black = 0:1,
  hispanic = 0:1,
  hgc_9 = 6:12 - 9,
  id = -999 # just use fixed effects
)
pred_frame <- pred_frame %>%
  mutate(pred = predict(m_wage1,
                        pred_frame,
                        allow.new.levels = TRUE))

# Race/Ethnicity
pred_frame <- pred_frame %>%
  mutate(
    race_eth = case_when(
      black == 0 & hispanic == 0 ~ "White",
      black == 1 & hispanic == 0 ~ "Black",
      black == 0 & hispanic == 1 ~ "Hispanic",
      TRUE ~ NA_character_
    )
  )

# Look at just hgc_9 == 0.
pred_frame %>%
  drop_na() %>%
  filter(hgc_9 == 0) %>%
  ggplot(aes(exper, pred)) +
  geom_line(aes(color = race_eth))
```

```{r interactions}
# If we want to know how the slope may or may not depend on these variables, we have to model the interactions.

m_wage2 <- lmer(lnw ~ exper + black + exper:black +
                  exper:hispanic + 
                  hgc_9 + exper:hgc_9 +
                  (exper|id),
                data = wages,
                control = lmerControl(optimizer = "bobyqa"))

# Make new predictions
pred_frame <- pred_frame %>%
  mutate(pred_int = predict(m_wage2,
                            newdata = pred_frame,
                            allow.new.levels = TRUE))

# Plot
pred_frame %>%
  drop_na() %>%
  filter(hgc_9 == 0) %>%
  ggplot(aes(exper, pred_int)) +
  geom_line(aes(color = race_eth))

pred_frame %>%
  drop_na() %>%
  filter(hgc_9 == -3 | hgc_9 == 3) %>%
  ggplot(aes(exper, pred_int)) +
  geom_line(aes(color = race_eth)) +
  facet_wrap(~hgc_9)

# Focus on hgc

pred_frame %>%
  drop_na() %>%
  ggplot(aes(exper, pred_int)) +
  geom_line(aes(color = factor(hgc_9))) +
  facet_wrap(~race_eth) +
  scale_color_brewer("Highest grade completed",
                     palette = "Accent",
                     breaks = 3:-3,
                     labels = 12:6) +
  labs(x = "Experience (years)",
       y = "Model Predicted wages (log scaled)")

# Model summary
arm::display(m_wage2)
```

```{r the_data_complexities}
sim_d %>%
  count(sid) %>%
  summarize(range(n))

# Varied "starting" points
sim_d %>%
  arrange(sid, date) %>%
  group_by(sid) %>%
  slice(1) %>%
  ungroup() %>%
  summarize(range(date))

# Overall date range
range(sim_d$date)
```

```{r plot_date_score}
ggplot(sim_d, aes(date, score)) +
  geom_point(alpha = 0.15, stroke = NA) +
  geom_smooth(se = FALSE, color = "#33B1AE", size = 2)
# Ideas on how to model this?
# Linear modeling is not going to work...
ggplot(sim_d, aes(date, score)) +
  geom_point(alpha = 0.15, stroke = NA) +
  geom_smooth(se = FALSE, color = "#33B1AE", size = 2) +
  geom_smooth(se = FALSE, method = "lm", color = "#808AFF", size = 2)
```

# Fit a model
Let's try fitting a linear model and a quadratic model and see which fits better. You try fitting the linear model first, with date predicting score, and both the intercept and slope varying across students.
## Center date
Let's first center date and put it in interpretable units.
I'll center it on the first time point. First - what do dates look like when converted to numbers?
```{r date_stuff}
# One unit = one day.
as_date(0)
as_date(1)
as_date(470)
# Center
sim_d <- sim_d %>%
  mutate(
    days_from_start = as.numeric(date) - min(as.numeric(date))
  )

# Fit linear model
linear <- lmer(score ~ days_from_start + (days_from_start|sid), 
               data = sim_d,
               control = lmerControl(optimizer = "Nelder_Mead"))
arm::display(linear)

# Fit quadratic model
sim_d <- sim_d %>%
  mutate(days2 = days_from_start^2)
quad <- lmer(score ~ days_from_start + days2 + 
               (days_from_start|sid), 
             data = sim_d,
             control = lmerControl(optimizer = "Nelder_Mead"))
arm::display(quad)

# Compare
anova(linear, quad)

# Plot predictions
pred_frame <- tibble(
    days_from_start = 0:max(sim_d$days_from_start),
    days2 = days_from_start^2,
    sid = -999
  ) %>%
  mutate(pred_linear = predict(linear, newdata = ., allow.new.levels = TRUE),
         pred_quad = predict(quad, newdata = ., allow.new.levels = TRUE))
pred_frame

ggplot(pred_frame, aes(days_from_start)) +
  geom_point(aes(y = score), data = sim_d, color = "gray80") +
  geom_line(aes(y = pred_linear), color = "#33B1AE") +
  geom_line(aes(y = pred_quad), color = "#808AFF")

# This is definitely looking better, but it's too high in the lower tail and maybe a bit too low in the upper.
# Interpret as "one this day, at this time... score equals"
```

```{r cubic}
sim_d <- sim_d %>%
  mutate(days3 = days_from_start^3)
cubic <- lmer(score ~ days_from_start + days2 + days3 +
                (days_from_start|sid), 
             data = sim_d,
             control = lmerControl(optimizer = "Nelder_Mead"))

arm::display(cubic)

# Compare
anova(linear, quad, cubic)

# Predictions
pred_frame <- pred_frame %>%
  mutate(days3 = days_from_start^3)
pred_frame %>%
  mutate(pred_cubic = predict(cubic, newdata = ., allow.new.levels = TRUE)) %>%
  ggplot(aes(days_from_start)) +
  geom_point(aes(y = score), data = sim_d, color = "gray80") +
  geom_line(aes(y = pred_linear), color = "#33B1AE") +
  geom_line(aes(y = pred_quad), color = "#808AFF") +
  geom_line(aes(y = pred_cubic), color = "#ff66fa")
```

```{r try_log}
sim_d <- sim_d %>%
  mutate(days_log = log(days_from_start + 1))
log_m <- lmer(score ~ days_log + (days_log|sid),
              data = sim_d)
arm::display(log_m)

# Compare
anova(linear, quad, cubic, log_m)
# It use the same number of parameters as the linear model, but fits far better.

# Predictions
pred_frame <- pred_frame %>%
  mutate(days_log = log(days_from_start + 1))
pred_frame <- pred_frame %>%
  mutate(pred_log = predict(log_m, newdata = ., allow.new.levels = TRUE))

# Predictions on log scale
ggplot(pred_frame, aes(days_log)) +
  geom_point(aes(y = score), data = sim_d, color = "gray80") +
  geom_line(aes(y = pred_log), color = "#4D4F57")

# Predictions on raw scale
pred_frame %>%
  mutate(pred_cubic = predict(cubic, newdata = ., allow.new.levels = TRUE)) %>%
  ggplot(aes(days_from_start)) +
  geom_point(aes(y = score), data = sim_d, color = "gray80") +
  geom_line(aes(y = pred_linear), color = "#33B1AE") +
  geom_line(aes(y = pred_quad), color = "#808AFF") +
  geom_line(aes(y = pred_cubic), color = "#ff66fa") +
  geom_line(aes(y = pred_log), color = "#4D4F57")
```